The Goldilocks Level of Friction: Using User Numbers, Ratings, and Reviews to Evaluate Digital Self-Control Tools

### 2022-04-29 paper.Rmd
incorporate initial proofing

The constant connectivity of smartphones and laptops provides innumerable benefits, but also unprecedented opportunities for distraction.
Indeed, as the business models of many tech companies have come to rely on nudging people into using digital services extensively (cf. the 'attention economy' [@davenportAttentionEconomyUnderstanding2001; @einsteinBlackOpsAdvertising2016; @wuAttentionMerchantsEpic2016]), many people routinely struggle to exercise self-control over their device use [@coxDesignFrictionsMindful2016; @koNUGUGroupbasedIntervention2015; @marottaOnlineDistractionsWebsite2017; @tranModelingEngagementDisengagementCycle2019; @hinikerMyTimeDesigningEvaluating2016].
As a result, public debate about negative effects of 'persuasive technologies' has been growing [@OptimizingEngagementUnderstanding2019; @orlowski2020] alongside increased political pressure on technology platforms to safeguard users [@allpartyparliamentarygrouponsocialmediaandyoungpeoplesmentalhealthandwellbeingNewFiltersManageImpact2019].

On online stores, a flourishing market exists for 'digital self-control tools', where hundreds of apps and browser extensions promise to help people take back control over their device use.
These tools offer interventions such as blocking apps or removing distracting website elements, tracking and visualising usage, or providing rewards or punishments for how devices are used [@lyngsSelfControlCyberspaceApplying2019].
Similarly, Apple and Google now offer various usage logging and self-limiting features in their 'ScreenTime' and 'Digital Wellbeing' tools, respectively [@appleIOS12Introduces2018; @gonzalezResearchGoogleNew2018; @googleDigitalWellbeing2018], and even Facebook have introduced tools 'for managing your time on Facebook and Instagram' [@ranadiveNewToolsManage2018].

In academia, a growing amount of HCI research is focusing on developing and evaluating various design patterns meant to help users self-regulate digital device use [@lottridgeBrowserDesignImpacts2012; @kimGoalKeeperExploringInteraction2019; @pinderPushAwaySmartphone2019; @tsengOvercomingDistractionsTransitions2019; @mongeroffarelloRaceDigitalWellbeing2019; @lyngsSelfControlCyberspaceApplying2019][^1].
This area of HCI work has gradually been building an understanding of how a range of design patterns from visualisations of time spent [@whittakerDonWasteMy2016] to goal-setting with social support [@koNUGUGroupbasedIntervention2015] can help people stay in control.
However, we are still in the early days of answering basic questions including how to define and measure 'success' for these design patterns, how much people vary in how useful they find particular interventions, and how specific design patterns' usefulness depends on details of their implementation, such as the level of 'friction' they introduce [@kimGoalKeeperExploringInteraction2019; @markEffectsIndividualDifferences2018; @tseng2019; @lyngsTellMeWhat2018; @hinikerMovingGreatScreen2019].

[^1]: We use the term 'design patterns' to refer to specific functions intended to support people in self-regulating their digital device use (e.g., hiding newsfeeds, providing goal reminders), and the term 'digital self-control tool' (DSCT) to refer to apps and browser extensions that implement a given design pattern (e.g., a browser extension for hiding Facebook's newsfeed).
    We occasionally use the term 'intervention' to refer to the implementation of a design pattern in the context of it changing a user's usual digital environment, or when it is evaluated in a research study.

To answer such questions, most HCI studies have taken an experimental approach and compared user behaviour and perceptions when applying a given intervention to its absence and/or some alternative intervention [@lyngs2021].
Yet, while controlled studies are appropriate and feasible for evaluating and comparing a small number of design patterns, they are difficult to scale to broadly assess a large numbers of patterns and implementations.

A complementary approach, which may help scope the range of design patterns and implementations to explore in controlled studies, is to investigate the existing landscape of digital self-control tools on app and web stores [@lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018].
Here, user reviews and popularity metrics can provide information about the relative utility of different patterns and implementations, contexts of use, and user needs 'in the wild'.
However, few studies to date have taken this approach.
Hence, in the present paper, we contribute to the growing HCI literature on design patterns for digital self-control by using app and web store information as a resource for understanding user self-experimentation with digital self-control tools (DSCTs).
Specifically, we present an analysis of user reviews, functionality, installation numbers, and ratings for 334 digital self-control tools, drawn from Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] who reviewed functionality in such tools on the Google Play, Chrome Web, and Apple App stores.

Our analysis of user reviews finds that digital self-control tools are seen as crucial to support focus on less instantly rewarding goals when digital distractions are easily available, specifically in contexts related to productivity (including work, studying, reading, and writing), and going to sleep.
DSCTs seem especially useful for people who describe themselves as 'addicts' or as struggling with distraction due to conditions like ADHD.

Users want DSCTs to flexibly adapt to their personal definition of 'distraction', as seen in common requests for customisation of *what* and *when* particular usage patterns should be blocked or tracked.
A key reason that users' and tools' conceptualisation of distraction should be in close alignment is that DSCTs' interventions, useful as they may be, can elicit negative emotional reactions.
This may cause the effects of interventions to be counterproductive, especially when users perceive DSCTs' goals or framing to differ from their own.

Finally, users want DSCTs to provide carefully calibrated incentives that are sufficient to incentivise intended behaviour change without feeling overly intrusive.
In our analysis, this 'Goldilocks level'[^2] varied between reviewers, but they commonly wanted tools to provide incentives powerful enough to keep themselves on track in moments of temptation or absentmindedness.
Overly strong incentives could backfire if they simply made users feel like a failure (in the case of reward/punishment tools) or did not allow some way to handle genuine emergency use (in the case of block/removal tools).

@@

One way for tools to provide properly calibrated incentives might be by carefully combining different types of design patterns (e.g., using both block/removal and goal-advancement features).
This was often asked for in reviews' feature suggestions, and supported by higher overall ratings given to tools combining two or more types compared to one.
A final consideration related to the amount of effort involved in operating a tool, where low friction to set up and initiate use (e.g., via schedules where a tool automatically turns on), but high friction to override it, was important.

@@

Other areas of HCI research have benefited from taking such an approach.
For example, in mental health research, thematic analysis of user reviews has been used to explore factors important for optimising user experience and support engagement in apps for cognitive behavioural therapy [@bakkerMentalHealthSmartphone2016; @huguetSystematicReviewCognitive2016; @luiEvidencebasedAppsReview2017; @sucalaAnxietyThereApp2017; @stawarzUserExperienceCognitive2018].
Similar research found that content in popular apps marketed as reducing symptoms of anxiety and worry was largely inconsistent with evidence-based treatments [@kertzReviewFreeIPhone2017].
This opened avenues for further research including how established CBT practices could more effectively be disseminated and, conversely, what might be learned from alternative approaches that proved popular in the wild.
HCI research on design patterns for digital self-control might in a similar way derive actionable insights from the large number of tools for this purpose available online.

However, few studies on digital self-control have so far taken this route: To the best of our knowledge, four published studies have described and/or categorised functionality in samples of digital self-control tools available online [@biskjaerTakingActionDistraction2016; @lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018]: Velthoven et al. [@vanvelthovenProblematicSmartphoneUse2018] presented aims and features in 21 tools identified via the software recommendation platform [alternativeto.net](https://alternativeto.net); Biskjær et al. [@biskjaerTakingActionDistraction2016] presented a taxonomy of functionality based on 10 tools from the Google Play, Chrome Web, and Apple App stores, and online tech magazines; Roffarello & Russis [@mongeroffarelloRaceDigitalWellbeing2019] presented a taxonomy of functionality based on 42 mobile apps from the Google Play store; and Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] presented a taxonomy of functionality based on 367 tools from the Google Play, Chrome Web and Apple App stores, which was mapped to psychological mechanisms in a dual systems model.

Only one of these studies analysed user reviews, namely Roffarello & Russis [@mongeroffarelloRaceDigitalWellbeing2019], which included thematic analysis of 1,128 reviews from 42 smartphone apps, and none incorporated popularity metrics.
Moreover, Roffarello & Russis' analysis of user reviews was restricted to 42 smartphone apps drawn from the Google Play store and did not specify how the reviews were distributed across apps with different functionality.
Thematic analysis is likely to yield different results depending on how reviews are sampled across tools with different functionality, and depending on the analysis strategy [@chenARminerMiningInformative2014; @braunThematicAnalysis2018].
Therefore, additional analyses with broader sampling of tools, and transparent sampling of reviews from those tools, might be helpful.

On this background, we extended existing studies of publicly available digital self-control tools by (i) scraping installation numbers, ratings, and user reviews from a broad sample of tools across stores and types of functionality, (ii) analysing review content using 'reflexive' thematic analysis [@braunThematicAnalysis2018], (iii) analysing popularity metrics' relationship to tools' design features, and (iv) making our materials, analysis pipelines and data openly available for the purpose of transparency and reusability [@harozUpdatesCHISubmission2019].

@@

Data, analysis scripts, and materials for the present paper are available at \url{https://osf.io/cdqub/}.

>>>>

The constant connectivity of smartphones and laptops provides innumerable benefits, but also unprecedented opportunities for distraction.
Indeed, as the business models of many tech companies have come to rely on nudging people into using digital services extensively (cf. the 'attention economy' [@davenportAttentionEconomyUnderstanding2001; @einsteinBlackOpsAdvertising2016; @wuAttentionMerchantsEpic2016]), many people routinely struggle to exercise self-control over their device use [@coxDesignFrictionsMindful2016; @koNUGUGroupbasedIntervention2015; @marottaOnlineDistractionsWebsite2017; @tranModelingEngagementDisengagementCycle2019; @hinikerMyTimeDesigningEvaluating2016].
As a result, public debate about negative effects of 'persuasive technologies' has grown [@OptimizingEngagementUnderstanding2019; @orlowski2020], alongside increased political pressure on technology platforms to safeguard users [@allpartyparliamentarygrouponsocialmediaandyoungpeoplesmentalhealthandwellbeingNewFiltersManageImpact2019].

On online stores, a flourishing market exists for 'digital self-control tools' (DSCTs), where hundreds of apps and browser extensions promise to help people take back control over their device use.
These tools offer interventions such as blocking apps or removing distracting website elements, tracking and visualising usage, or providing rewards or punishments for how devices are used [@lyngsSelfControlCyberspaceApplying2019].
Similarly, Apple and Google now offer various usage logging and self-limiting features in their 'ScreenTime' and 'Digital Wellbeing' tools, respectively [@appleIOS12Introduces2018; @gonzalezResearchGoogleNew2018; @googleDigitalWellbeing2018], and even Facebook have introduced tools 'for managing your time on Facebook and Instagram' [@ranadiveNewToolsManage2018].

In academia, a growing amount of HCI research is similarly focused on developing and evaluating various design patterns that help users self-regulate digital device use [@lottridgeBrowserDesignImpacts2012; @kimGoalKeeperExploringInteraction2019; @pinderPushAwaySmartphone2019; @tsengOvercomingDistractionsTransitions2019; @mongeroffarelloRaceDigitalWellbeing2019; @lyngsSelfControlCyberspaceApplying2019][^1].
This work has gradually built an understanding of how a range of design patterns from visualisations of time spent [@whittakerDonWasteMy2016] to goal-setting with social support [@koNUGUGroupbasedIntervention2015] can help people stay in control.
However, we are still in the early days of answering basic questions including how to define and measure 'success' for these design patterns, how implementation details of specific design patterns --- such as the level of 'friction' --- influence their usefulness, and how much people vary in how useful they find particular interventions [@kimGoalKeeperExploringInteraction2019; @markEffectsIndividualDifferences2018; @tseng2019; @lyngsTellMeWhat2018; @hinikerMovingGreatScreen2019].

[^1]: We use the term 'design patterns' to refer to specific functions intended to support people in self-regulating their digital device use (e.g., hiding newsfeeds, providing goal reminders).
    We use the term 'digital self-control tool' (DSCT) to refer to apps and browser extensions that implement a given design pattern (e.g., a browser extension for hiding Facebook's newsfeed).
    We occasionally use the term 'intervention' to refer to the implementation of a design pattern in the context of it changing a user's usual digital environment, or when it is being evaluated in a research study.

To answer such questions, most existing studies have taken an experimental approach and compared user behaviour and perceptions when applying a given intervention to its absence and/or some alternative intervention [@lyngs2021].
Yet, while controlled studies are appropriate and feasible for evaluating and comparing a small number of design patterns, they are difficult to scale to broadly assess a large numbers of patterns and implementations.

A complementary approach, which can help scope the range of design patterns and implementations to explore in controlled studies, is to investigate the existing landscape of DSCTs on app and web stores [@lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018].
Here, user reviews and popularity metrics can provide information about user needs, contexts of use, and the relative utility of different patterns and implementations 'in the wild'.
However, few studies to date have taken this approach.
Hence, in the present paper, we contribute to the HCI literature on design patterns for digital self-control by using app and web store information as a resource for understanding user self-experimentation with DSCTs.
Specifically, we present an analysis of user reviews, functionality, installation numbers, and ratings for 334 digital self-control tools, drawn from Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] who reviewed functionality in such tools on the Google Play, Chrome Web, and Apple App stores.

Our analysis of user reviews finds that digital self-control tools are seen as highly important to supports focus on less instantly rewarding goals when digital distractions are easily available, specifically in contexts related to productivity (including work, studying, reading, and writing), and going to sleep.
DSCTs appeared to be especially useful for people who describe themselves as 'addicts' or as struggling with distraction due to conditions like ADHD.

Users want DSCTs to adapt to their personal definition of 'distraction', as seen in common requests for customisation of *what* and *when* particular usage patterns should be blocked or tracked.
One important reason that users' and tools' conceptualisation of distraction should be in close alignment is that DSCTs' interventions, useful as they may be, can elicit negative emotional reactions.
This may cause the effects of interventions to be counterproductive, especially if users perceive DSCTs' goals or framing to differ from their own.

Finally, users want DSCTs to provide carefully calibrated incentives that are sufficient to incentivise intended behaviour change without feeling overly intrusive.
In our analysis, DSCTs 'Goldilocks level'[^2] varied between reviewers.
However, reviewers commonly wanted tools that provided incentives strong enough to keep themselves on track in moments of temptation or absentmindedness.
Overly strong incentives could, however, backfire if they made users feel like a failure (in the case of reward/punishment tools) or did not allow some way to handle genuine emergency use (in the case of block/removal tools).

@@

One way for tools to provide properly calibrated incentives might be by carefully combining different types of design patterns (e.g., using both block/removal and goal-advancement features).
This was often asked for in reviews' feature suggestions, and tentatively supported by higher overall ratings given to tools that combine two or more types compared to just one.
A final consideration related to the amount of effort involved in operating a tool, where low friction to set up and initiate use (e.g., via schedules where a tool automatically turns on), but high friction to override it, was important.

@@

Other areas of HCI research have benefited from taking such an approach.
For example, in mental health research, thematic analysis of user reviews has been used to explore factors important for optimising user experience and support engagement in apps for cognitive behavioural therapy [@bakkerMentalHealthSmartphone2016; @huguetSystematicReviewCognitive2016; @luiEvidencebasedAppsReview2017; @sucalaAnxietyThereApp2017; @stawarzUserExperienceCognitive2018].
<!-- Similar research found that content in popular apps marketed as reducing symptoms of anxiety and worry was largely inconsistent with evidence-based treatments [@kertzReviewFreeIPhone2017]. --> <!-- This opened avenues for further research including how established CBT practices could more effectively be disseminated and, conversely, what might be learned from alternative approaches that proved popular in the wild. --> <!-- HCI research on design patterns for digital self-control might in a similar way derive actionable insights from the large number of tools for this purpose available online. --> However, few studies on digital self-control have so far taken this route: To the best of our knowledge, four published studies have described and/or categorised functionality in samples of digital self-control tools available online [@biskjaerTakingActionDistraction2016; @lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018]: Velthoven et al. [@vanvelthovenProblematicSmartphoneUse2018] presented aims and features in 21 tools identified via the software recommendation platform [alternativeto.net](https://alternativeto.net); Biskjær et al. [@biskjaerTakingActionDistraction2016] presented a taxonomy of functionality based on 10 tools from the Google Play, Chrome Web, and Apple App stores, and online tech magazines; Roffarello & Russis [@mongeroffarelloRaceDigitalWellbeing2019] presented a taxonomy of functionality based on 42 mobile apps from the Google Play store; and Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] presented a taxonomy of functionality based on 367 tools from the Google Play, Chrome Web and Apple App stores, which was mapped to psychological mechanisms in a dual systems model.

Only one of these studies analysed user reviews, namely Roffarello & Russis [@mongeroffarelloRaceDigitalWellbeing2019], which included thematic analysis of 1,128 reviews from 42 smartphone apps, and none incorporated popularity metrics.
Moreover, Roffarello & Russis' analysis was restricted to 42 smartphone apps drawn from the Google Play store and did not specify how the reviews were distributed across apps with different functionality.
Thematic analysis is likely to yield different results depending on how reviews are sampled across tools with different functionality, and depending on the analysis strategy [@chenARminerMiningInformative2014; @braunThematicAnalysis2018].
Therefore, additional analyses with broader sampling of tools, and transparent sampling of reviews from those tools, might be helpful.

On this background, we extended existing studies of publicly available digital self-control tools by (i) scraping installation numbers, ratings, and user reviews from a broad sample of tools across stores and types of functionality, (ii) analysing review content using 'reflexive' thematic analysis [@braunThematicAnalysis2018], and (iii) analysing popularity metrics' relationship to tools' design features, and (iv) making our materials, analysis pipelines and data openly available for the purpose of transparency and reusability [@harozUpdatesCHISubmission2019].

@@

Data, analysis scripts, and materials for the present paper are available at [osf.io/cdqub/](https://osf.io/cdqub/).
